Low Complexity Adaptive Sonar Imaging

### 2016-05-01 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2016:05:01	20:45:42	(sø. 1.mai)

To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

@@

\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough. This is why human made sonars have little to no adaptivity on transmit. However, several methods have been investigated to adaptively post process the received data. Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected. However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them. This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR. How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.

@@

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular. Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards. Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$. Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.

Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Section \ref{sec:conclusion} provides a conclusion.

@@

MVDR seeks to minimize the output variance of the beamformer, under the constraint of a unity gain in some desired direction $\phi$~\cite{Capon1969}. With the assumption of zero mean data this can be formulated as: 
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
\end{gather}
The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent diagonal loading~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/buske1_online.eps}%
\else%
\includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
\fi%
\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\mathrm{3\,dB}}(\beta)]$. The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
\end{figure}

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/buske2_online.png}%
\else%
\includegraphics[width=\linewidth]{gfx/buske2_online.png}%
\fi%
\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups for the shadow, highlight and speckle region of the image. The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale. The size of the pixel groups were chosen ad-hoc for the histograms to be near invariant to a shift in position within the same region. Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside. The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight. The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
\end{figure*}

\begin{figure*}[t]%
\includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied. The upper left image is equal to DAS with a rectangular window. For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object). These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$. \Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
\emph{Upper bound $\phi$:}\hfill
\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

\begin{figure*}[t]\centering
\subfloat[$\beta$-values of the Kaiser window chosen for each image pixel. Note how LCA prefers a a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference. On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/buske5a_online.pdf}%%
\label{oversampling_mosaic_bounds_beta}}
\hfil
\subfloat[$\phi$-values of the Kaiser window chosen for each image pixel. On strong sources LCA selects windows that are steered away from the source. This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/buske5b_online.pdf}%
\label{oversampling_mosaic_bounds_phi}}
\caption{Kaiser windows chosen for each pixel in the image. The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters. The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
\label{fig_sim}
\end{figure*}

\begin{figure*}[t]%\centering%
\includegraphics[width=\textwidth]{gfx/buske6_online.pdf}%
\caption{\emph{LCA images using different window database sizes}. The window boundaries are the same. The upper left image is identical to a rectangularly weighted DAS. Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged. Adding more steering-variations improves both sidelobe suppression and resolution. However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
\end{figure*}

@@

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criterion:
%
\begin{align}
\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq:mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights. As will be demonstrated in section \ref{sec:results_lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

@@

If the pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this. However, we observed no visual improvement from applying this technique, and decided not to use it.
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window. A Kaiser window $\f_\beta$ steered to the angle $\phi$ can be expressed as 
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
}.\label{eq:steering_vector}
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

Each window's response is constrained to unit gain in the look direction. Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}. To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones. We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window. The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.

@@

While we focus on using LCA with Kaiser windows, we will for completeness compare the performance with that of using trigonometric windows. This is defined as:
%
\begin{align}
f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\label{eq:trig_window_function}
\end{align}
%
where we apply the contraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients. Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images are all bilinearly upinterpolated to the same size. 
% 
% \todo{Need $M$ in the image sector. Why???}

>>>>

To build confidence in the LCA method we show that a robust MVDR implementation typically creates weight sets with shapes spanning between a rectangular and Hamming window function. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

@@

\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough. This is why human made sonars have little to no adaptivity on transmit. However, several methods have been investigated to adaptively process the received acustic data. Perhaps the best studied method for reconstructing a sonar image out of such data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA)~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows is selected. However, LCA differs in that it typically selects from a larger pool of windows and allows phase-steered variations of them. Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards. This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR. How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.

@@

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he let LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular. Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While he achieved good results, it was not made clear how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$. Although this system was designed as a synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.

Our results show that when MVDR is robustified to work in active sonar imaging, it computes weights with spatial amplitude responses that tend to be symmetric and steered within a fraction of their \minus{}3\,dB widths. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the \minus{}3\,dB width of each window's spatial amplitude response. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

This article is outlined as follows: In Section \ref{sec:beamforming} we offer a gentle introduction to beamforming. Then we move on to adaptive beamforming and describe the MVDR and LCA methods in \ref{sec:mvdr} and \ref{sec:lca}, respectively. We use LCA with the Kaiser window which we describe in \ref{sec:lca_kaiser_windows}, explain how we steer it in \ref{sec:lca_steering}, and by how much in \ref{sec:lca_steering_bounds}. Finally we add some remarks on using the trigonometric window instead of Kaiser in \ref{sec:lca_trigonometric}, and of of sampling considerations in \ref{sec:lca_oversampling}. Results and discussion are provided in Section \ref{sec:results_discussion}, where we study MVDR's windows to find suitable LCA ones in \ref{sec:results_lca_window_function}, assess the range of window types and steering in \ref{sec:results_window_parameters}, determine the number of windows needed in \ref{sec:results_database_size}, and add a note on computational complexity in \ref{sec:results_complexity}. Finally, Section \ref{sec:conclusion} provides a conclusion.

@@

MVDR seeks to minimize the power of noise and interference in the output of the beamformer, under the constraint of unity gain in some desired direction $\phi$~\cite{Capon1969}. With the assumption of zero mean data this takes the form of minimizing output variance: 
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w^H[n]\a_\phi &= 1.\label{eq:mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector for an azimuthal steering angle $\phi$, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. This is a convex optimization problem with the unique solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi}.\label{eq:mvdr_weights}
\end{gather}
The problem lies in estimating and inverting the spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}: To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}, for true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}, and to improve robustness to parameter errors we add {\large$\epsilon$} percent of the total output power to the matrix's diagonal~\cite{Cox1987,Maksym1979}. These steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/buske1_online.eps}%
\else%
\includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
\fi%
\caption{Each Kaiser window is steered in the interval $\phi\in[0, \phi_{\mathrm{3\,dB}}(\beta)]$. The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered amplitude response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
\end{figure}

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/buske2_online.png}%
\else%
\includegraphics[width=\linewidth]{gfx/buske2_online.png}%
\fi%
\caption{The submerged object used to test beamforming resolution is a 1\,m by 1\,m test cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with 40x40 pixel groups for the shadow, highlight and speckle region of the image indicated with red boxes.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from the 40x40 pixel groups. The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale. The size of the pixel groups were chosen ad-hoc for the histograms to be visually invariant to a shift in position within the same region. Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside. The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight. The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
\end{figure*}

\begin{figure*}[t]%
\includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large window database, but where the upper bounds for $\beta$ and $\phi$ are varied. The upper left image is equal to DAS with a rectangular window. For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object). These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$. \Fig{mvdr_selected_windows} suggests that MVDR prefers windows with sidelobe levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
\emph{Upper bound $\phi$:}\hfill
\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

\begin{figure*}[t]\centering
\subfloat[$\beta$-values of the Kaiser window chosen for each image pixel. Note how LCA prefers a narrow response ($\beta=0$) on the anchor, and on other strong sources with little lateral interference. On the sources where lateral interference is present it chooses the widest response with the best sidelobe suppression.]{\includegraphics[width=\textwidth]{gfx/buske5a_online.pdf}%%
\label{oversampling_mosaic_bounds_beta}}
\hfil
\subfloat[$\phi$-values of the Kaiser window chosen for each image pixel. On strong sources LCA selects windows that are steered away from the source. This is what improves the FWHM measurement in \Fig{oversampling_mosaic_bounds}.]{\includegraphics[width=\textwidth]{gfx/buske5b_online.pdf}%
\label{oversampling_mosaic_bounds_phi}}
\caption{Kaiser windows chosen for each pixel in the image of the resolution test cross. The underlying image is the one shown in \Fig{oversampling_mosaic_bounds}, with the same parameters. The location of the anchor, cut line and main scatter locations of the cross is marked in red.}
\label{fig_sim}
\end{figure*}

\begin{figure*}[t]%\centering%
\includegraphics[width=\textwidth]{gfx/buske6_online.pdf}%
\caption{\emph{LCA images using different window database sizes}. The window boundaries are the same. The upper left image is identical to a rectangularly weighted DAS. Observe that finer sampling of the $\beta$-range improves sidelobe suppression but leaves resolution unchanged. Adding more steering-variations improves both sidelobe suppression and resolution. However, using more than  $N_\beta=2$ window types and $N_\phi=3$ steering angles makes minimal difference.}\label{oversampling_mosaic}
\end{figure*}

@@

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criterion:
%
\begin{align}
\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq:lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq:mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solution space for the weights. As will be demonstrated in Section \ref{sec:results_lca_window_function}, LCA performs similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

@@

If the pixels are correlated laterally, we can include a weighted combination of these to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this. However, we observed no visual improvement from applying this technique, and decided not to use it.
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

Adding slightly steered versions of each window to the window database gives LCA greater flexibility in searching for an optimal window. A Kaiser window $\f_\beta$ steered to the azimuth angle $\phi$ can be expressed as 
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\label{eq:steered_window}
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{2\pi d}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{2\pi (M-1)d}{\lambda}\sin(\phi)}
}.\label{eq:steering_vector}
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq:lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

Each window's spatial response is constrained to unit gain in the look direction. Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}. To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones. We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered amplitude response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window. The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.

@@

While we focus on using LCA with the Kaiser window function, we will for comparison also test it with the trigonometric window function. Its definition is:
%
\begin{align}
f_m(\alpha) = \alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big),\label{eq:trig_window_function}
\end{align}
%
where we apply the constraint $\alpha\in[0.5,1]$ to avoid windows with negative coefficients. Steering is applied as in (\ref{eq:steered_window}), and the $\alpha$-value that halves the \minus{}3\,dB distance is found using the method described in Appendix \ref{appendix:steering_angles}.

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images were all bilinearly upinterpolated to the same size. 
% 
% \todo{Need $M$ in the image sector. Why???}
